---
title: "assignment_3"
format: html
aurthor: Lauren Chetley & Emma Rodriguez
editor: visual
---

## Assignment 3

### Overview

This assignment involves simulating one year of daily air quality monitoring data, classifying pollution risk levels, and analyzing seasonal patterns in air quality conditions. All work should be completed within this Quarto document with all code and corresponding output clearly displayed.

## Part 1

```{r}

set.seed(123)

# create day column and generate realistic PM2.5 values
air_quality <- data.frame(
  day = 1:365,
  pm25 = rnorm(365, mean = 18, sd = 12)
)

# create season column
air_quality$season <- factor(
  ifelse(air_quality$day <= 90, "winter",
         ifelse(air_quality$day <= 180, "spring",
                ifelse(air_quality$day <= 270, "summer", "fall"))),
  levels = c("winter", "spring", "summer", "fall")
)

# create empty risk column
air_quality$risk <- NA

head(air_quality)
```

## Part 2

```{r}

# create empty risk column
air_quality$risk <- NA

# variables to track unhealthy streaks
current_unhealthy_streak <- 0
max_unhealthy_streak <- 0

# for loop to classify risk levels
for(i in 1:nrow(air_quality))
  {
  if(air_quality$pm25[i] < 12)  # good air quality
    { air_quality$risk[i] <- "good"
      current_unhealthy_streak <- 0}
  
  else if(air_quality$pm25[i] >= 12 & air_quality$pm25[i] <= 35.4)  # moderate air quality 
    { air_quality$risk[i] <- "moderate"
      current_unhealthy_streak <- 0}
  
  else {air_quality$risk[i] <- "unhealthy" # unhealthy air quality 
  
  if(air_quality$risk[i] == "unhealthy")
  {current_unhealthy_streak <- current_unhealthy_streak + 1
  
  if(current_unhealthy_streak > max_unhealthy_streak)
  {max_unhealthy_streak <- current_unhealthy_streak}}
  
  else
  {current_unhealthy_streak <- 0}}
}

head(air_quality)

# longest unhealthy streak 
max_unhealthy_streak

summary(air_quality$pm25)
```

## Part 3

```{r}

# create vector to store unhealthy day counts
unhealthy_days <- c(winter = 0, spring = 0, summer = 0, fall = 0)

for(i in 1:nrow(air_quality)) 
  {
  
  if(air_quality$risk[i] == "unhealthy") 
    {season_name <- as.character(air_quality$season[i])
    unhealthy_days[season_name] <- unhealthy_days[season_name] + 1}
}

unhealthy_days
```

## Part 4

### Plot 1: Histogram showing PM2.5 concentrations across the year

```{r}
library(tidyverse)
library(ggplot2)

#| fig-cap: "Figure 1. Distribution of simulated PM2.5 concentrations by season."

hist_concentration <- ggplot(air_quality, aes(x = pm25, fill = season)) +
  geom_histogram() +
   scale_fill_manual(values = c(
     "winter" = "lightblue",
    "spring" = "palegreen2",
    "summer" = "lightgoldenrod1",
    "fall" = "sienna3")) +
  labs(title = sprintf("PM2.5 Concentrations by Season, Max Unhealthy Streak: %s Days", max_unhealthy_streak),
    x = "PM2.5 Concentration",
    y = "Number of Days")+
    theme_minimal()

hist_concentration
```

## Interpretation of PM2.5 Monitoring Results 

The simulated air quality data showed seasonal differences in PM2.5 and unhealthy air quality events throughout the year. Summer experienced the highest number of unhealthy days with 10 total days classified as unhealthy. Winter had the fewest unhealthy days at 6. Spring and fall showed moderate levels of unhealthy air quality with 7 and 8 days. The maximum consecutive streak of unhealthy air was only 2. Overall PM2.5 concentration had and average value of 18.38 ug/m3, with some extreme pollution events reaching a max of 56.89 ug/m3. These findings suggest that while most days experienced moderate air quality conditions, occasional shirt term pollution spikes could still pose risks to public health.
